Ages of Stars and Planets in the Kepler Field Over the First Four Billion Years

### 2024-08-29 papers/paper/ms.tex
update ms

% upon AAS submission
\documentclass[11pt,twocolumn,tighten,linenumbers]{aastex63}
%\documentclass[11pt,twocolumn,tighten,linenumbers,trackchanges]{aastex63}
% drafting / arxiv
%\documentclass[11pt,twocolumn,tighten]{aastex63}
\turnoffedit

@@

% <250 words
\begin{abstract}
  %
  Recent analyses of FGK stars in open clusters have helped clarify
  the precision with which a star's rotation rate and lithium content
  can be used as empirical indicators for its age.
  %
  Here we apply this knowledge to stars observed by Kepler.
  %
  Rotation periods are drawn from previous work; lithium is measured
  from new and archival Keck/HIRES spectra.
  %
  We report rotation-based ages for \nuniqstarsantosrotgyroappl\ stars
  (harboring \nplwgyroagenograzing\ known planets) for which our method is
  applicable.
  %
  We find that our rotational ages accurately recover the ages of
  stars in open clusters spanning 0.04-2.5\,Gyr; they also agree with
  $\gtrsim$90\% of the independent lithium ages.
  %
  The resulting yield includes \nplyounggyrotwosigma\ planets younger
  than 1\,Gyr at 2$\sigma$, and \nplyounggyro\ with median ages below
  1\,Gyr.
  %
  This is about half the number expected under the classic assumption
  of a uniform star formation history.  We find that the scarcity of
  sub-gigayear systems can be attributed to the star formation rate in
  the Kepler field \replaced{dropping}{smoothly decreasing} by a factor of \replaced{\ratiosfr$\pm$\uncratiosfr}{3-5}
  over the past 3\,Gyr.  We observe this trend both for known planet
  hosts and for the parent stellar sample.
  %
  This ``demographic cliff'' in the Galaxy's star formation history
  has been previously reported, and its confirmation helps clarify the
  age distribution of the known transiting exoplanets.
  %
\end{abstract}

@@

Young ($<$1\,Gyr) exoplanets represent one means of building the
timeline for exoplanet evolution.  Informative individual exemplars
include HIP~67522b, a Jupiter-sized planet with a sub-Neptunian mass
(\citealt{Rizzuto_2020}; Thao et al.~submitted), and V1298~Tau, a
resonant chain of puffy planets that is a likely precusor to the
compact multiplanet systems \citep{David_2019}.  Population-level
analyses have similarly suggested differences in the size distribution
of young exoplanets relative to their older counterparts
\citep{Berger_2020b_rpage,David_2021,Sandoval_2021,2023AJ....166..248C,2024arXiv240303261V}.

@@

\begin{enumerate}[label={\it \roman*)},leftmargin=12pt,topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
  \item The Kepler Object of Interest catalog, and our vetting of
    false positives within it, has now reached maturity
    \citep[e.g.][]{Thompson_2018}.
  %
  \item Measured rotation periods of FGK stars in open clusters now
    show not only the average evolution of $P_{\rm rot}(T_{\rm
    eff},t)$, but also how the astrophysical dispersion of stars
    around this average converges by the $\approx$700\,Myr age of
    Praesepe
    \citep[e.g.][]{Curtis_2019_ngc6811,Gillen_2020,Rampalli_2021,Fritzewski_2021,Rebull_2022,Dungee_2022,2023AJ....166...14B}.
  %
  \item Using open cluster data, we can marginalize over the range of
    ages that might explain any one star's rotation period
    \citep{Bouma_2023}.  This represents an improvement in the
    accuracy of uncertainty propagation relative to previous
    calibrations.
  %
  \item We can identify binary stars \added{and high-metallicity outliers }with greater fidelity than
    in the past, which can clarify otherwise problematic estimates of
    rotation-based ages.
  %
  \item New open clusters in various stages of dissolution have been
    found in the Kepler field
    \cite[e.g.][]{2019AJ....158..122K,Kerr_2021,Barber_2022}.
    These clusters offer stars that we can use to test
    the reliability of the available single-star age-dating methods.
\end{enumerate}

@@

\added{{\it Completeness}---Figure~\ref{fig:hist_tgyro} has a gray
overlay beyond $\gtrsim$3\,Gyr because Kepler's sensitivity to
rotation signals diminishes at old ages.} \citet{2022ApJ...937...94M}
studied \deleted{this issue}\added{Kepler's completeness to rotation
signals using MIST-derived isochrone ages}, and found for Sun-like
stars that the fraction of rotation signals \added{that are
}detectable \deleted{by Kepler }is near unity up to $\approx$3\,Gyr,
and that it drops to almost zero by $\approx$5\,Gyr.  \deleted{This trend is
opposite in functional form to our derived age
distributions.}\added{This estimate suggests that while aspects of
Figure~\ref{fig:hist_tgyro} might be interpretable in terms of the Galaxy's star
formation history at $\lesssim$3\,Gyr, at older ages incompleteness
likely precludes any such interpretation.}

@@

Sub-gigayear planets can be particularly informative for studies of
planet evolution.  Our catalog has \nplyounggyro\ confirmed and
candidate planets with median ages below 1\,Gyr.  Requiring $t_{\rm
gyro}$$<$1\,Gyr at 2$\sigma$ yields \nplyounggyrotwosigma\ planets
orbiting \nplhostsyounggyrotwosigma\ stars.  The youth claim is most
secure for the systems that are either in clusters (Section~\ref{subsec:clusterplanets}), or that have
independent rotation and lithium-based ages (Section~\ref{subsec:rotnlithiumcomp}).
These samples provide broader context for the planets around field
stars that push the boundaries of the current young exoplanet census
(Section~\ref{subsec:notables}).

@@

Among these systems, four highlights are Kepler-1529, Kepler-1565,
Kepler-1312, and Kepler-1629.  Kepler-1529b is a
$\approx$100$\pm$50\,Myr mini-Neptune with well-measured rotation and
lithium ages.  Kepler-1565b is an analogous $\approx$170-230\,Myr
super-Earth.  Kepler-1312 is a $t_{\rm gyro}$=\kepthirteentwelve\,Myr
near-Solar analog with an Earth-sized planet on a one-day orbit, and a
mini-Neptune on a five-day orbit.  The Earth-sized planet,
Kepler-1312~c, along with the near-identical Kepler-1561~b ($t_{\rm
gyro}$=\kepfifteensixone\,Myr), rank among the youngest Earth-sized
planets currently known, comparable to planets such as HD~63433d
\citep[1.1\,$R_\oplus$, $414$$\pm$$23$\,Myr;][]{2024AJ....167...54C}
and TOI-1807~b \citep[1.3\,$R_\oplus$,
180$\pm$40\,Myr;][]{2021AJ....162...54H}.  Kepler-1629~b, with a size
just two-thirds that of Earth (0.67$\pm$0.05\,$R_\oplus$), is
marginally older, also based on rotation (\kepsixteentwonine\,Myr).
The lithium age from a reconnaissance TRES spectrum confirms the
point.

@@

\subsection{The Thin Disk's Star Formation History}

Most \deleted{of the }stars observed by Kepler are in the thin disk.  This can
be verified by following \citet{Gaia_2018}, and labeling stars with 2D
tangential velocities $v_{\rm T}$$<$40\,\kms\ as thin disk members,
and those with 60$<$$v_{\rm T}$$<$150\,\kms\ as thick disk members.
\replaced{For all stars in the KIC, this yields }{These criteria yield }95{,}694 thin and 54{,}039 thick
disk stars \replaced{respectively}{observed by Kepler}.  The thin disk stars have a threefold larger
detected rotation period fraction: 35{,}675 rotators are in the thin
disk, while 7{,}312 are in the thick disk.  Further imposing the
\deleted{gyrochronology }quality flags discussed in Section~\ref{subsec:flags}
yields 17{,}755 thin disk stars for which gyrochronology is
applicable, and 2{,}462 thick disk stars.

Classifying Kepler stars as thin vs.\ thick disk members helps connect
them to previous work on the Galaxy's star formation history.  On
extragalactic scales, the star formation rate in spiral galaxies with
mass similar to the Milky Way peaked $\approx$10\,Gyr ago, and has
since decreased by an order of magnitude
\citep[e.g.][]{2004Natur.428..625H,2006ApJ...651..142H}.  In our own
galaxy, previous studies have measured star formation histories
through CMD fitting of resolved stellar populations
\citep[][]{2019A&A...624L...1M,2020NatAs...4..965R,2021MNRAS.501..302A,2022Natur.603..599X}
and by modeling the local white dwarf luminosity function
\citep[e.g.][]{2019ApJ...878L..11I}.  \citet{2019A&A...624L...1M} and
\citet{2019ApJ...878L..11I} for instance focused on stars (and white
dwarfs) within a few hundred parsecs, and both found a peak in the
local star formation rate 2-3\,Gyr ago.  \citet{2020NatAs...4..965R}
focused on stars in a wider 2\,kpc bubble, and additionally reported
three local maxima in the galactic star formation rate, which they
associated with pericenter passages of the Sagittarius satellite
galaxy.

The right column of Figure~\ref{fig:hist_tgyro} compares our derived
age distribution against star formation histories (SFHs) reported by
\citet{2019A&A...624L...1M} and \citet{2020NatAs...4..965R} based on
CMD fitting.  The overall slope of both SFHs broadly agrees with what
we find from rotation-based ages.  The SFH from
\citet{2019A&A...624L...1M} seems entirely consistent, particularly
after accounting for the statistical uncertainties of that study.
Regarding the episodic star formation bursts reported by
\citet{2020NatAs...4..965R}, they are not apparent in our overall FGK
star sample (top row).  However, rotation-based ages are most precise
for $\approx$G8V-K4V dwarfs older than $\approx$1.5\,Gyr
\citep{Bouma_2023}.  The lower row of Figure~\ref{fig:hist_tgyro}
selects these stars with a temperature cut, and may show a \added{weak }hint of the
\added{$\approx$1\,Gyr and }$\approx$2\,Gyr spike\added{s} reported by \citet{2020NatAs...4..965R}.  \deleted{The
$\approx$1\,Gyr spike however is not recovered. } This
$\approx$2.5\,Gyr local maximum is similarly present in the isochrone
ages derived by \citet{Berger_2020a_catalog} for the Kepler field, and
in the red giant asteroseismic ages derived by
\citet{SilvaAguirre2018}.  The main novelty of our ages relative to
\replaced{those studies is their improved accuracy}{previous work is that they are demonstrably accurate} at $<$1\,Gyr\added{ (see Figure~\ref{fig:agescalecompone})}.

@@

Our most constraining ages tend to come from one source of
information: rotation.  Factors other than \deleted{stellar }age and mass can
influence \deleted{stellar }rotation rates.  \added{Astrophysical factors include
metallicity, differential rotation, mergers, and tidal spin-up.  Observational
limitations include the difficulty of measuring small-amplitude,
long-period signals that are only quasi-coherent due to the starspot
lifetime approaching the stellar rotation period.  There is also the issue that
our age-dating approach is not calibrated beyond 4\,Gyr.
In the following, we attempt to
quantify the importance of these effects in our analysis, ordered by
what we perceive to be their descending importance.}

@@

The sizes and periods for the most secure set of planets are shown in
Figure~\ref{fig:rp_period_age_results}.   While the new young planets
are mostly mini-Neptunes, some are near the lower boundary of the
``sub-Jovian desert'' \citep{Owen2018}, which could have an
evolutionary connection to their youth.  Other discoveries include
Earth-sized planets \added{(e.g.~Kepler-1312~c and Kepler-1561~b) }with \deleted{new }ages of only a few hundred million years.

@@

\begin{enumerate}[leftmargin=*,topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
  %
  \item Rather than being uniform, the age distributions of both the
    Kepler target stars and the known Kepler planet hosts show a
    demographic cliff.  There are \added{at least }twice as many ``old'' (2-3\,Gyr)
    stars in the Kepler field as ``young'' (0-1\,Gyr) stars.  The star
    formation rate today is\added{ at least} \ratiosfr$\pm$\uncratiosfr\ times lower
    than it was three billion years ago\added{.  The statistical
    uncertainty on this value is dwarfed by the systematic
    uncertainty caused by the potential presence of unresolved
    photometric binaries (Section~\ref{subsec:caveats}); the star
    formation rate could have decreased by a factor of as much as
    4.81$\pm$0.26 over this time interval}.  \replaced{This}{Either} result \replaced{from
    rotation-based ages broadly agrees}{is generally consistent} with recent reports of a
    declining star-formation rate from CMD fitting and white-dwarf
    chronology, though with the \added{potential }advantage of better \replaced{sensitivity}{accuracy} for
    FGK stars at $t$$<$1\,Gyr (e.g.~Figure~\ref{fig:agescalecompone}).
  %
  \item Rather than expecting $\approx$500 exoplanets younger than one
    billion years, this age distribution of stars in the Kepler field
    \replaced{implies}{suggests} that we should instead expect $\approx$250.
  %
  \item We have derived rotation-based ages for \nplyounggyrotwosigma\
    Kepler planets younger than 1\,Gyr at 2$\sigma$, and
    \nplyounggyro\ planets with median ages below 1\,Gyr.
    Concatenating these planets against the existing literature yields
    $\approx$170 known sub-gigayear planets.  This lessens the
    original factor of ten discrepancy to a factor of at most two.
\end{enumerate}

\acknowledgements
We thank the reviewer for their suggestions, which significantly improved the 
quality of the manuscript.
This work was supported by the Heising-Simons 51~Pegasi~b Fellowship
(LGB, EKP) and the Arthur R.~Adams SURF Fellowship (EKP).
The HIRES data were obtained at the Keck Observatory.
We recognize the importance that the summit of Maunakea has always had
within the indigenous Hawaiian community, and we are deeply grateful 
for the opportunity to conduct observations from this mountain.

>>>>

% upon AAS submission
%\documentclass[11pt,twocolumn,tighten,linenumbers]{aastex63}
%\documentclass[11pt,twocolumn,tighten,linenumbers,trackchanges]{aastex63}
% drafting / arxiv
\documentclass[11pt,twocolumn,tighten]{aastex63}
\turnoffedit

@@

% <250 words
\begin{abstract}
  %
  Recent analyses of FGK stars in open clusters have helped clarify
  the precision with which a star's rotation rate and lithium content
  can be used as empirical indicators for its age.
  %
  Here we apply this knowledge to stars observed by Kepler.
  %
  Rotation periods are drawn from previous work; lithium is measured
  from new and archival Keck/HIRES spectra.
  %
  We report rotation-based ages for \nuniqstarsantosrotgyroappl\ stars
  (harboring \nplwgyroagenograzing\ known planets) for which our method is
  applicable.
  %
  We find that our rotational ages accurately recover the ages of
  stars in open clusters spanning 0.04-2.5\,Gyr; they also agree with
  $\gtrsim$90\% of the independent lithium ages.
  %
  The resulting yield includes \nplyounggyrotwosigma\ planets younger
  than 1\,Gyr at 2$\sigma$, and \nplyounggyro\ with median ages below
  1\,Gyr.
  %
  This is about half the number expected under the classic assumption
  of a uniform star formation history.
  %
  \deleted{We find that the scarcity of sub-gigayear systems can be
  attributed to the star formation rate in the Kepler field dropping
  by a factor of \ratiosfr$\pm$\uncratiosfr over the past 3\,Gyr.  We
  observe this trend both for known planet hosts and for the parent
  stellar sample.  This ``demographic cliff'' in the Galaxy's star
  formation history has been previously reported, and its confirmation
  helps clarify the age distribution of the known transiting
  exoplanets.}
  %
  \added{The age distribution that we observe, rather than being uniform,
  shows that the youngest stars in the Kepler field are 3-5 times
  rarer than stars 3\,Gyr old.
  This trend holds for both known planet hosts and for the parent
  stellar sample.
  We attribute the existence of this ``demographic cliff'' to a
  combination of kinematic heating and a declining star formation rate
  in the Galaxy's thin disk, and highlight its impact on the age
  distribution of known transiting exoplanets.
  }
\end{abstract}

@@

Young ($<$1\,Gyr) exoplanets represent one means of building the
timeline for exoplanet evolution.  Informative individual exemplars
include HIP~67522\,b, a Jupiter-sized planet with a sub-Neptunian mass
(\citealt{Rizzuto_2020}; Thao et al.~submitted), and V1298~Tau, a
resonant chain of puffy planets that is a likely precusor to the
compact multiplanet systems \citep{David_2019}.  Population-level
analyses have similarly suggested differences in the size distribution
of young exoplanets relative to their older counterparts
\citep{Berger_2020b_rpage,David_2021,Sandoval_2021,2023AJ....166..248C,2024arXiv240303261V}.

@@

\begin{enumerate}[label={\it \roman*)},leftmargin=12pt,topsep=0pt,itemsep=-1ex,partopsep=1ex,parsep=1ex]
  \item The Kepler Object of Interest catalog, and our vetting of
    false positives within it, has now reached maturity
    \citep[e.g.][]{Thompson_2018}.
  %
  \item Measured rotation periods of FGK stars in open clusters now
    show not only the average evolution of $P_{\rm rot}(T_{\rm
    eff},t)$, but also how the astrophysical dispersion of stars
    around this average converges by the $\approx$700\,Myr age of
    Praesepe
    \citep[e.g.][]{Curtis_2019_ngc6811,Gillen_2020,Rampalli_2021,Fritzewski_2021,Rebull_2022,Dungee_2022,2023AJ....166...14B}.
  %
  \item Using open cluster data, we can marginalize over the range of
    ages that might explain any one star's rotation period
    \citep{Bouma_2023}.  This represents an improvement in the
    accuracy of uncertainty propagation relative to previous
    calibrations.
  %
  \item We can identify \replaced{binary stars}{stars that are
    binaries and that have high metallicity} with greater fidelity\deleted{ than
    in the past}, which can clarify otherwise problematic estimates of
    rotation-based ages.
  %
  \item New open clusters in various stages of dissolution have been
    found in the Kepler field
    \cite[e.g.][]{2019AJ....158..122K,Kerr_2021,Barber_2022}.
    These clusters offer stars that we can use to test
    the reliability of the available single-star age-dating methods.
\end{enumerate}

@@

\added{{\it Completeness}---Figure~\ref{fig:hist_tgyro} has a gray
overlay beyond $\gtrsim$3\,Gyr because Kepler's sensitivity to
rotation signals diminishes at old ages.} \citet{2022ApJ...937...94M}
studied \deleted{this issue}\added{Kepler's completeness to rotation
signals using MIST-derived isochrone ages}, and found for Sun-like
stars that the fraction of rotation signals \added{that are
}detectable \deleted{by Kepler }is near unity up to $\approx$3\,Gyr,
and that it drops to almost zero by $\approx$5\,Gyr.  \deleted{This trend is
opposite in functional form to our derived age
distributions.}\added{This estimate suggests that while aspects of
Figure~\ref{fig:hist_tgyro} might be interpretable in terms of the Galaxy's star
formation history at $\lesssim$3\,Gyr, at older ages incompleteness
precludes any such interpretation.}

@@

\deleted{Sub-gigayear planets
can be particularly informative for studies of planet
evolution.}
\added{This analysis provides an important path toward expanding the population
of young and evolving exoplanets.}
Our catalog has \nplyounggyro\ confirmed
and candidate planets with median ages below 1\,Gyr.  Requiring
$t_{\rm gyro}$$<$1\,Gyr at 2$\sigma$ yields \nplyounggyrotwosigma\
planets orbiting \nplhostsyounggyrotwosigma\ stars.  The youth claim
is most secure for the systems that are either in clusters
(Section~\ref{subsec:clusterplanets}), or that have independent
rotation and lithium-based ages
(Section~\ref{subsec:rotnlithiumcomp}).  These samples provide broader
context for the planets around field stars that push the boundaries of
the current young exoplanet census (Section~\ref{subsec:notables}).

@@

Among these systems, four highlights are Kepler-1529, Kepler-1565,
Kepler-1312, and Kepler-1629.  Kepler-1529\,b is a
$\approx$100$\pm$50\,Myr mini-Neptune with well-measured rotation and
lithium ages.  Kepler-1565\,b is an analogous $\approx$170-230\,Myr
super-Earth.  Kepler-1312 is a $t_{\rm gyro}$=\kepthirteentwelve\,Myr
near-Solar analog with an Earth-sized planet on a one-day orbit, and a
mini-Neptune on a five-day orbit.  The Earth-sized planet,
Kepler-1312\,c, along with the near-identical Kepler-1561\,b ($t_{\rm
gyro}$=\kepfifteensixone\,Myr), rank among the youngest Earth-sized
planets currently known, comparable to planets such as HD~63433\,d
\citep[1.1\,$R_\oplus$, $414$$\pm$$23$\,Myr;][]{2024AJ....167...54C}
and TOI-1807\,b \citep[1.3\,$R_\oplus$,
180$\pm$40\,Myr;][]{2021AJ....162...54H}.  Kepler-1629\,b, with a size
just two-thirds that of Earth (0.67$\pm$0.05\,$R_\oplus$), is
marginally older, also based on rotation (\kepsixteentwonine\,Myr).
The lithium age from a reconnaissance TRES spectrum confirms the
point.

@@

\begin{figure*}[!t]
	\begin{center}
		\leavevmode
		\subfloat{
			\includegraphics[width=.99\textwidth]{threepanel_trilegal_comparison_twostepsfr.pdf}
		}
	\end{center}
	\vspace{-0.5cm}
	\caption{{\bf Impact of star formation rate (SFR) and kinematic heating on age distribution.}   
		Most Kepler stars are a few hundred parsecs above the Galactic plane ($\hat{Z}$ points to the north galactic pole;  $\hat{Y}$ is the direction of
		galactic rotation).  This is comparable to the $\approx$300\,pc mass-weighted scale height.
		The dotted gray and solid orange lines show synthetic age distributions
		from TRILEGAL assuming a constant SFR (middle) and a two-step
		SFR (right), described in the text.
		A combination of both kinematic heating and
		a decreasing SFR qualitatively match the observed paucity of the youngest stars.
		\label{fig:trilegal}
	}
\end{figure*}

%\subsection{The Thin Disk's Star Formation History}
\subsection{Kinematic Heating and Star Formation in the Thin Disk}
\label{subsec:trilegal}

\added{Two separate astrophysical effects might explain our observed 
age distribution (Figure~\ref{fig:hist_tgyro}): a declining star formation
rate (SFR), and kinematic heating.   Kinematic heating is associated with an
increase in stellar velocity dispersions and the disk's vertical 
scale height over time \citep[e.g.][]{Villumsen1983,Schmidt2024}.
The Kepler field points toward $(l,b)$=(76.3$^\circ$,13.5$^\circ$).
A typical Kepler star at 1\,kpc is therefore $\approx$230\,pc
above the galactic plane, a significant fraction of the thin disk's 
mass-weighted $\approx$300\,pc scale height \citep{Bland-Hawthorn2016}.
}

\added{We assessed the relative importance of these two effects
	using the TRILEGAL Monte Carlo population synthesis tool }\citep[v1.6;][]{Girardi2005}.
\added{TRILEGAL treats the Galaxy as a linear combination of a thin disk, thick disk,
	halo, and bulge, and allows for two possible SFRs: constant, and a ``two-step''
	SFR in which the thin disk SFR is 50\% larger from 1-4\,Gyr than at other ages.
	This two-step SFR was introduced to improve the model's ability to
	match star counts from Hipparcos and 2MASS }\citep[see Section~5.1 of][]{Girardi2005}.
\added{For kinematic heating, TRILEGAL assumes that
	the disk scale height grows as
$h_{\rm z} = z_0 (1 + t/t_0)^\alpha$, for $\alpha$=5/3, $z_0$=94.7\,pc,
	and $t_0$=5.55\,Gyr, similar to \citet{Rana1991}.
	We queried the web interface toward the
	Kepler field, and toward the galactic plane (at $l$=76.3$^\circ$) to
	negate the effects of kinematic heating.
	We limited the output to stars with apparent Kepler magnitudes brighter than 16,
	and set the field of view to 10 ${\rm deg}^2$, the maximum
	possible through the web interface.
%	and set the Sun at $z_\odot$=20.8\,pc, $r$=8122\,pc from the Galactic center.
}

\added{Figure~\ref{fig:trilegal} compares the simulation results against
	our observed age distribution.
	The middle panel shows that assuming a constant SFR,
	kinematic heating alone can yield a two-fold difference
	in the abundance of the youngest stars relative to those $\approx$3\,Gyr old.
	The right-most panel
	shows that additionally including 
	the time-variable SFR yields the best agreement with the slope of
	our observed age distribution.
%	An important note is that the previously noted parameters
%	that were fine-tuned to fit different datasets, primarily photometric star counts
%	along various sight-lines.  The exercise appears to have been predictive.
}

\deleted{Most stars observed by Kepler are in the thin disk.
This can be verified by following \citet{Gaia_2018}, and labeling
stars with 2D tangential velocities $v_{\rm T}$$<$40\,\kms\ as thin
disk members, and those with 60$<$$v_{\rm T}$$<$150\,\kms\ as thick
disk members.  These
criteria yield 95{,}694 thin and 54{,}039 thick disk stars
observed by Kepler.  The thin disk stars have
a threefold larger detected rotation period fraction: 35{,}675
rotators are in the thin disk, while 7{,}312 are in the thick disk.
Further imposing the quality flags discussed
in Section~\ref{subsec:flags} yields 17{,}755 thin disk stars for
which gyrochronology is applicable, and 2{,}462 thick disk stars.}

\deleted{Classifying Kepler stars as thin vs.\ thick disk members helps connect
them to previous work on the Galaxy's star formation history.  On
extragalactic scales, the star formation rate in spiral galaxies with
mass similar to the Milky Way peaked $\approx$10\,Gyr ago, and has
since decreased by an order of magnitude
\citep[e.g.][]{2004Natur.428..625H,2006ApJ...651..142H}.}
\added{Despite the expected differences between pencil-beam surveys
and volume-complete samples, our observed age distribution qualitatively
agrees with a number of previous studies.}
\replaced{In our own
galaxy, previous studies have}{Previous work has} measured star formation histories
through CMD fitting of resolved stellar populations
\citep[][]{2019A&A...624L...1M,2020NatAs...4..965R,2021MNRAS.501..302A,2022Natur.603..599X},
and by modeling the local white dwarf luminosity function
\citep[e.g.][]{2019ApJ...878L..11I}.  \citet{2019A&A...624L...1M} and
\citet{2019ApJ...878L..11I} for instance focused on stars (and white
dwarfs) within a few hundred parsecs, and both found a peak in the
local star formation rate 2-3\,Gyr ago.  \citet{2020NatAs...4..965R}
focused on stars in a wider 2\,kpc bubble, and additionally reported
three local maxima in the SFR, which they
associated with pericenter passages of the Sagittarius satellite
galaxy.

The right column of Figure~\ref{fig:hist_tgyro} compares our derived
age distribution against star formation histories (SFHs) reported by
\citet{2019A&A...624L...1M} and \citet{2020NatAs...4..965R} based on
CMD fitting.  The overall slope of both SFHs broadly agrees with what
we find from rotation-based ages.  The SFH from
\citet{2019A&A...624L...1M} seems entirely consistent, particularly
after accounting for the statistical uncertainties of that study.
Regarding the episodic star formation bursts reported by
\citet{2020NatAs...4..965R}, they are not apparent in our overall FGK
star sample (top row).  However, rotation-based ages are most precise
for $\approx$G8V-K4V dwarfs older than $\approx$1.5\,Gyr
\citep{Bouma_2023}.  The lower row of Figure~\ref{fig:hist_tgyro}
selects these stars with a temperature cut, and may show a \added{weak }hint of the
\added{$\approx$1\,Gyr and }$\approx$2\,Gyr spike\added{s} reported by \citet{2020NatAs...4..965R}.  \deleted{The
$\approx$1\,Gyr spike however is not recovered. } This
$\approx$2.5\,Gyr local maximum is similarly present in the isochrone
ages derived by \citet{Berger_2020a_catalog} for the Kepler field, and
in the red giant asteroseismic ages derived by
\citet{SilvaAguirre2018}.  The main novelty of our ages relative to
\replaced{those studies is their improved accuracy}{these previous studies is their accuracy} at $<$1\,Gyr\added{ (see Figure~\ref{fig:agescalecompone})}.

@@

Our most constraining ages tend to come from one source of
information: rotation.  Factors other than \deleted{stellar }age and mass can
influence \deleted{stellar }rotation rates.  \added{Astrophysical factors include
metallicity, latitudinal differential rotation, mergers, and tidal spin-up.  Observational
limitations include the difficulty of measuring small-amplitude,
long-period signals that are only quasi-coherent due to the starspot
lifetime approaching the stellar rotation period.  There is also the issue that
our age-dating approach is not calibrated beyond 4\,Gyr.
In the following, we attempt to
quantify the importance of these effects in our analysis, ordered by
what we perceive to be their descending importance.}

@@

The sizes and periods for the most secure set of planets are shown in
Figure~\ref{fig:rp_period_age_results}.   While the new young planets
are mostly mini-Neptunes, some are near the lower boundary of the
``sub-Jovian desert'' \citep{Owen2018}, which could have an
evolutionary connection to their youth.  Other discoveries include
Earth-sized planets \added{(e.g.~Kepler-1312\,c and Kepler-1561\,b) }with \deleted{new }ages of only a few hundred million years.

@@

\begin{enumerate}[leftmargin=*,topsep=0pt,itemsep=-0.5ex,partopsep=1ex,parsep=1ex]
  %
  \item Rather than being uniform, the age distributions of both the
    Kepler target stars and the known Kepler planet hosts show a
    demographic cliff.  There are \added{at least }twice as many ``old'' (2-3\,Gyr)
    stars in the Kepler field as ``young'' (0-1\,Gyr) stars.  \deleted{The star
    formation rate today is\added{ at least} \ratiosfr$\pm$\uncratiosfr\ times lower
    than it was three billion years ago.}
\item \added{The youngest stars in the Kepler field (0-0.3\,Gyr)
    are at least  \ratiosfr$\pm$\uncratiosfr\ times rarer than stars 3\,Gyr old. The statistical
    uncertainty on this value is dwarfed by the systematic
    uncertainty in the contamination rate of unresolved
    photometric binaries, which could push it as high as 4.81$\pm$0.26 (Section~\ref{subsec:caveats})}.
     \replaced{This}{Either} result \replaced{from
    rotation-based ages broadly agrees}{is broadly consistent} with \replaced{recent reports of a
    declining star-formation rate}{modeling expectations (Section~\ref{subsec:trilegal}), and with age distributions derived from} from CMD fitting and white-dwarf
    chronology\replaced{, though with the}{.  The }advantage of \added{rotation-based ages however is their }\replaced{sensitivity}{accuracy} for
    FGK stars at $t$$<$1\,Gyr (e.g.~Figure~\ref{fig:agescalecompone}).
  %
  \item Rather than expecting $\approx$500 exoplanets younger than one
    billion years, the age distribution of stars in the Kepler field
    \replaced{implies}{suggests} that we should instead expect $\approx$250.
  %
  \item We have derived rotation-based ages for \nplyounggyrotwosigma\
    Kepler planets younger than 1\,Gyr at 2$\sigma$, and
    \nplyounggyro\ planets with median ages below 1\,Gyr.
    Concatenating these planets against the existing literature yields
    $\approx$170 known sub-gigayear planets.  This lessens the
    original factor of ten discrepancy to a factor of at most two.
\end{enumerate}

\acknowledgements
We thank Kevin Schlaufman and the reviewer for suggestions which significantly improved the 
manuscript.
This work was supported by the Heising-Simons 51~Pegasi~b Fellowship
(LGB, EKP) and the Arthur R.~Adams SURF Fellowship (EKP).
The HIRES data were obtained at the Keck Observatory.
We recognize the importance that the summit of Maunakea has always had
within the indigenous Hawaiian community, and we are deeply grateful 
for the opportunity to conduct observations from this mountain.
